Measuring contrast processing in the visual system using SSVEP (and VEP?)

### 2024-03-27 review.qmd
Added ,some motivation

In early work on infants, the ‘sweep VEP’ paradigm was developed to measure SSVEP activity in response to a stimulus that changes its properties throughout a trial, for example gradually increasing in contrast or spatial frequency. Other studies keep the stimulus properties constant for the entire trial, which avoids serial effects and makes the analysis more straightforward.

Because it provides a direct read-out of neural population activity, the SSVEP signal can reveal key features of neural signal transduction. By parametrically varying the peak stimulus contrast, a monotonically increasing contrast response function can be measured. This corresponds closely to similar functions measured in single unit studies, or those measuring local field potentials invasively in the cortex. However SSVEP has the advantage that it is non-invasive, and so can be measured in awake, behaving human participants. An early use of the contrast response function was to provide an automated estimate of contrast sensitivity, without requiring behavioural responses from the participant. This was achieved by measuring the contrast response function, and extrapolating back along the function to estimate its intercept with the x-axis, which was shown to correspond with psychophysically measured detection thresholds (REFS). This general approach revealed much about the development of visual abilities in infants (Braddick), as well as deficits in neurological disease. However, psychophysical measurements remain the gold standard, and do not always correspond closely with SSVEP estimates in individuals with normal adult visual systems.

>>>>

### Why measure responses to contrast?
Contrast is one of the most fundamental pieces of information that the eye transmits to the brain. It can be defined as the change in cone photoreceptor activity over time ('spatial contrast') or space ('spatial contrast'). Cone photoreceptors contribute to both chromatic and achromatic contrast and although most of the research we describe here focuses on achromatic contrast, SSVEPs have proven to be an excellent measure of early chromatic processing as well. [McKeefry, Baseler, DiRusso...].

Contrast is relatively simple to define: typically, contrast is specificed as the percentage deviation of uniform stimulus from the background. So, for example, a disk of 100 units, $I_{\mathrm{stim}}$ of cone activation surrounded by a 'background' of 50 units of activation $I_{\mathrm{background}}$ has a contrast of $\frac{I_{\mathrm{stim}} - I_{\mathrm{background}}}{I_{\mathrm{background}}}$ = 100%. Where patterns are more complex (for example, the sine-wave gratings or Gabors common in vision science), the Michaelson definition of contrast is specified by the maximum and minimum excursions from the mean: $${\displaystyle {\frac {I_{\mathrm {stimmax} }-I_{\mathrm {stimmin} }}{I_{\mathrm {stimmax} }+I_{\mathrm {stimmin} }}},} $$. These contrast definitions are appropriate both to photometric measures of stimulus contrast (for example, luminance [LenniePokornySmith]) and also to definitions based on cone excitations [DKL, MacBoynton] which are more common in work on chromatic processing.

Although its definition is simple, the computations that underlie contrast processing have been the subject of intense research for many decades. The neural code for contrast, even in the earliest parts of visual cortex, is not simply a linear transform of the contrast at the retina - instead, contrast signals undergo a cascade of nonlinear processing stages that, broadly, attempt to normalise the output relative to the spatiotemporal environment. This normalization, achieved through a computation called 'contrast gain control' [Heeger 92, Heeger and Carandini 2011] maximises the sensitivity of the visual system by making optimal use of neuronal bandwidth. As an example, a grating seen on a low-contrast background tpyically appears more intense than the same grating when superimposed on a high contrast background. A signifcant part of research into contrast processing is concerned with how these normalization mechanisms depend on colour, orientation, eye of origin, spatial and temporal frequency, age, and the presence of neurological disorders. The SSVEP has proven to be invaliable in this research because it provides an objective readout of contrast representation at different parts of the visual system and allows us to 'tag' the probe and background separately.

### The development of contrast processing
In early work on infants, the ‘sweep VEP’ paradigm was developed to measure SSVEP activity in response to a stimulus that changes its properties throughout a trial, for example gradually increasing in contrast or spatial frequency. Other studies keep the stimulus properties constant for the entire trial, which avoids serial effects and makes the analysis more straightforward.

Because it provides a direct read-out of neural population activity, the SSVEP signal can reveal key features of neural signal transduction. By varying the peak stimulus contrast parametrically, a 'contrast vs response' function can be measured - where the 'response' is typically defined as the amplitude of the SSVEP frequency component at the stimulus frequency or a low multiple thereof. This corresponds closely to similar functions measured in single unit studies, or those measuring local field potentials invasively in the cortex. However SSVEP has the advantage that it is non-invasive, and so can be measured in awake, behaving human participants. An early use of the contrast response function was to provide an automated estimate of contrast sensitivity, without requiring behavioural responses from the participant. This was achieved by measuring the contrast response function, and extrapolating back along the function to estimate its intercept with the x-axis, which was shown to correspond with psychophysically measured detection thresholds (REFS). This general approach revealed much about the development of visual abilities in infants (Braddick), as well as deficits in neurological disease. In well-motivated adults, psychophysical measurements remain the gold standard. However, it is hard to obtain reliable psychophysical data from infants, children and adults with neuological disorders. In these cases, SSVEP measurements represent a fast and efficient method for measuring low-level visual responses.
